---
title: "The HistData Challenge: Re-Visioning Historical Graphics"
output:
  html_document:
    toc: true
    toc_float: true
    df_print: kable
---

<!--
WORKING DRAFT -- not a package vignette yet.

Lives in issues/vignettes/ deliberately, not vignettes/: pkgdown scans and publishes everything
under vignettes/ on every `pkgdown::build_site()`, so a half-finished draft there would go live
by accident. issues/ is already .Rbuildignore'd and outside pkgdown's scan path.

When this is ready to ship:
  1. Add a proper vignette YAML header (rmarkdown::html_vignette + %\VignetteIndexEntry etc.)
  2. Add a references.bib and switch citations below to [@key] form
  3. git mv issues/vignettes/histdata-challenge-draft.Rmd vignettes/histdata-challenge.Rmd
  4. devtools::document(); devtools::check(); NEWS.md entry; rebuild pkgdown
     (see issues/TASKS.md's "Workflow: rebuilding pkgdown" checklist)

See issues/vignettes/histdata-challenge.md for the full planning/rationale behind every
decision below -- this file is the draft prose+code, that one is the working notes.
-->

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "##",
  warning = FALSE,
  message = FALSE,
  fig.width = 7,
  fig.height = 5
)
library(HistData)
library(ggplot2)
```

> RE-VISION *n.* ri-'vizh-en (ca. 1611) 1. To see again, possibly from a new perspective; *syn:*
> review, reconsideration, reexamination, retrospection. 2. An act of revising; *syn:* rewrite,
> alteration, transformation. (Merriam-Webster, 2002, quoted in Friendly, 2002)

## Introduction

The `HistData` package collects data behind some of the most important graphics in the history
of statistics and data visualization. Some of these present graphical challenges to reproduce in
R, and this collection is part of a program of research called *statistical historiography*: the
use of statistical methods to study problems and questions in the history of statistics and
graphics. A main aspect of this is the increased understanding of historical problems in science
and data analysis through the process of trying to reproduce a graph or analysis using modern
methods. We call this "re-visioning" -- meaning to *see again, hopefully in a new light*.

This idea first took shape in Friendly (2002), "Visions and Re-Visions of Charles Joseph
Minard," which reviewed decades of attempts -- by many different people, in many different tools
-- to reproduce and extend Minard's famous graphic of Napoleon's Russian campaign. That paper,
and the gallery of re-visions that grew out of it, is the model for this vignette: a catalog of
*kinds* of re-vision, applied not just to Minard but to any dataset in this package, plus a
running inventory of what's already been attempted (in `sandbox/`, by outside contributors, in
companion projects) and what's still open.

## A small motivating example: `Arbuthnot`

Before anything as ambitious as Minard, a small example makes the point concretely. John
Arbuthnot's data on christenings in London (1629-1710) look, on a first read, like a straightforward
annual count. But Sandy Zabell (1976) found that the values for 1674 and 1704 are identical --
not approximately close, *identical*, down to a ratio computed to three decimal places -- strongly
suggesting one year's entry was copied from the other, whether by Arbuthnot himself or a later
transcriber.

```{r arbuthnot-duplicate}
data(Arbuthnot)
Arbuthnot[Arbuthnot$Year %in% c(1674, 1704), c("Year", "Males", "Females", "Ratio")]
```

`Males`, `Females`, and the derived `Ratio` match exactly; only `Plague` and `Mortality` (not
shown) differ. A table alone makes this easy to miss among 82 rows. Plotting the series makes it
harder to miss -- the two points sit at *exactly* the same height, 30 years apart, in a series
that otherwise wanders:

```{r arbuthnot-plot}
ggplot(Arbuthnot, aes(x = Year, y = Ratio)) +
  geom_line(color = "grey50") +
  geom_point(color = "grey50") +
  geom_point(
    data = subset(Arbuthnot, Year %in% c(1674, 1704)),
    color = "firebrick", size = 3
  ) +
  labs(
    title = "Arbuthnot's sex ratio at christening, 1629-1710",
    subtitle = "1674 and 1704 (red) are identical -- likely a copying error",
    y = "Male / female ratio"
  ) +
  theme_minimal()
```

That's re-visioning in miniature: looking again at old data, graphically, surfaces something
tabulation alone hides. See Zabell (1976) and Zabell & Wainer (2002) for the fuller story --
`?Arbuthnot` for the original citation, and `issues/TASKS.md` for a note on tracking down DOIs
for both.

## What is re-visioning?

*(Expand this section with more of the framing from Friendly (2002) and `?HistData` -- draft
placeholder for now; the Introduction above already covers the core definition.)*

## A taxonomy of re-visions

The gallery accompanying Friendly (2002), <https://www.datavis.ca/gallery/re-minard.php>,
catalogs about 15 different re-visions of the Minard graphic. Read as a checklist rather than a
one-off catalog, they sort into six recognizable *kinds* of re-vision -- a rubric that applies to
any HistData dataset, not just Minard's.

### Software/language ports

The same graphic, redrawn in a different tool: Mathematica, a formal Grammar-of-Graphics
specification, ggplot2, SAS/IML, Protovis. The content doesn't change; what changes is what the
choice of tool reveals about the graphic's underlying structure -- a grammar-of-graphics port, for
instance, forces you to name the mapping from data to visual channel explicitly, in a way the
original hand-drawn version never had to.

### Design variations

Same content, same tool family, different design decisions: legibility fixes, a graphic
designer's redraw, an alternate visual metaphor (Minard's gallery includes a pictograph version
emphasizing human cost over cartographic accuracy). This is the category where taste and
communication goals do the most work.

### Interactive/web-based

Tooltips, drill-down, linked views, map overlays -- re-visions that trade the fixed, printed page
for something a reader can query. The Minard gallery includes an image map with linked text and a
drill-down SAS GMAP; this is also the natural home for the `plotly` version of `Perozzo` below.

### Temporal/animated

A step-by-step reveal, a third axis given over to time, a narrated video. Minard's original
already compresses time into distance along the march route; an animated re-vision instead lets
time *play out*, trading the compression for a different kind of legibility. (See "Open
invitations" below for the deferred `gganimate` idea for `Perozzo`.)

### Alternative representations

A genuinely different chart type for the same underlying data -- not a redraw, a rethink. The
`Snow-density.R` case study below (density contours instead of point symbols for cholera deaths)
is this category in miniature.

### Data packaged for reuse

The original figures plus the underlying data, released in a form someone else can build on. This
is, quite literally, what the `HistData` package itself does for all 34 of its datasets --
arguably the least glamorous but most consequential category, since every other kind of re-vision
in this list depends on it having happened first.

*(TODO: pull a representative thumbnail image from the datavis.ca gallery for each category above
once hosting/licensing for third-party images is sorted out -- see `issues/vignettes/histdata-challenge.md`.)*

## Challenges already named in HistData

Several datasets in this package already carry an explicit "go recreate or improve this" note in
their documentation:

| Dataset | Nature of the challenge |
|---|---|
| [`Minard`](../reference/Minard.html) | "Minard's Challenge": reproduce the March on Moscow graphic with modern tools (Friendly, 2002). |
| [`Guerry`](../reference/Guerry.html) | Friendly (2007): "Challenges for Multivariable Spatial Analysis." |
| [`Playfair1824`](../reference/Playfair1824.html) | Re-create Playfair's last chart, or do better, using modern graphics. |
| [`Perozzo`](../reference/Perozzo.html) | Get closer to Perozzo's hand-drawn stereogram, or do better in some way -- the case study below. |
| [`Pollen`](../reference/Pollen.html) | The original 1986 ASA JSM "Data Challenge" -- a synthetic 5D dataset with several deliberately hidden features. |
| [`OldMaps`](../reference/OldMaps.html) | Open call to relate map accuracy to other characteristics of each map. |

This first draft works two of these (`Minard`, `Perozzo`) in depth, plus one further example from
outside the package's own docs (`Snow`, via John Russell's 30DayChartChallenge). The rest stay
listed here, cited, and open -- see "Open invitations."

## Case study: `Minard`

*(Draft placeholder. Per the plan: point to a couple of gallery entries from
<https://www.datavis.ca/gallery/re-minard.php>, one per taxonomy category above, as illustration
-- not a full reproduction. E.g. the ggplot2 port for "software/language ports," the pictograph
for "design variations," the animated GIF for "temporal/animated.")*

## Case study: `Perozzo`

Perozzo's 1880/1881 stereogram plots survivor counts as a surface over `Year` x `Age`. The
package ships a base-R version already:

```{r perozzo-setup}
data(Perozzo)
Pmat <- xtabs(Survivors ~ Year + Age, data = Perozzo)
years <- as.numeric(rownames(Pmat))
ages  <- as.numeric(colnames(Pmat))
```

```{r perozzo-base, fig.height = 5}
ages_rev <- -rev(ages)
Pmat_rev <- Pmat[, rev(seq_along(ages))]
persp(years, ages_rev, Pmat_rev,
      xlab = "Year", ylab = "Age", zlab = "Survivors",
      theta = 0, phi = 25, expand = 0.6,
      col = adjustcolor("lightblue", alpha.f = 0.5), shade = 0.5)
```

### Re-vision 1: ggplot2 raster + contour

A straightforward "alternative representation" re-vision: give up the 3D illusion entirely in
favor of a 2D heatmap with isolines, which trades the stereogram's dramatic silhouette for exact,
directly-comparable values.

```{r perozzo-ggplot-contour}
ggplot(Perozzo, aes(x = Year, y = Age, z = Survivors)) +
  geom_raster(aes(fill = Survivors)) +
  geom_contour(color = "white", alpha = 0.6) +
  scale_fill_distiller(palette = "Blues", direction = 1) +
  scale_y_reverse() +
  labs(title = "Perozzo's survivorship grid, as a heatmap") +
  theme_minimal()
```

### Re-vision 2: a fake-3D isometric shear

A "design variation": keep the 3D *look* Perozzo was after, but build it as an ordinary 2D ggplot
via a manual isometric shear (offsetting each `Age` slice horizontally and vertically as a
function of `Age`), rather than `persp()`'s true 3D projection. Unlike `persp()`, this gets full
access to ggplot's color scales, legends, and layering.

```{r perozzo-isometric}
shear <- 0.6   # horizontal points per year of age; tune to taste
lift  <- 400   # vertical Survivors-units per year of age

Perozzo_iso <- transform(Perozzo,
  x_iso = Year + shear * Age,
  y_iso = Survivors + lift * Age
)

ggplot(Perozzo_iso, aes(x = x_iso, y = y_iso, group = Age, color = Age)) +
  geom_line() +
  scale_color_distiller(palette = "Blues", direction = 1) +
  labs(
    title = "Perozzo's stereogram as a 2D isometric shear",
    x = "Year (sheared by Age)", y = "Survivors (lifted by Age)"
  ) +
  theme_minimal() +
  theme(axis.text.x = element_blank(), axis.text.y = element_blank())
```

*(Rough first pass -- the shear/lift constants above are not tuned, and this doesn't yet attempt
Perozzo's actual oblique axis convention. Revisit before shipping.)*

### Re-vision 3: an interactive `plotly` surface

An "interactive/web-based" re-vision, and arguably the most direct modern analogue of a physical
stereogram: something a reader can actually rotate, rather than a single fixed viewing angle.

```{r perozzo-plotly}
library(plotly)
plot_ly(x = ~years, y = ~ages, z = ~Pmat, type = "surface") |>
  layout(scene = list(
    xaxis = list(title = "Year"),
    yaxis = list(title = "Age"),
    zaxis = list(title = "Survivors")
  ))
```

### Still open

The trace-line coloring Perozzo himself used -- highlighting `age` and `year` cross-sections in
red at 25-year intervals (see `R/Perozzo.R`'s `@details`) -- was deferred earlier and isn't
attempted in any of the three re-visions above. Left as an open invitation (below) rather than
solved in this draft.

## Case study: `Snow`, via John Russell's 30DayChartChallenge

Of the `sandbox/` scripts recreating Russell's [30DayChartChallenge
2025](https://github.com/drjohnrussell/30DayChartChallenge) entries, `Snow-density.R` was picked
for this draft because it needs no dependencies beyond `ggplot2` (see
`issues/vignettes/histdata-challenge.md` for the full dependency audit of every candidate script),
and because it's a genuine "alternative representation" of one of the package's most iconic
datasets: density contours instead of point symbols for cholera deaths, layered over Snow's
original street map.

```{r snow-density}
data(Snow.deaths); data(Snow.pumps); data(Snow.streets)

ggplot(Snow.deaths, aes(x = x, y = y)) +
  geom_line(data = Snow.streets, aes(x = x, y = y, group = street),
            color = "black", linewidth = 0.5) +
  geom_point(alpha = 0.6) +
  geom_density_2d_filled(bins = 4, show.legend = FALSE) +
  geom_density_2d(bins = 4, color = "black", alpha = 0.6) +
  scale_fill_manual(values = c("#FFFFFF00", "#F0808030", "#FF000030", "#8B000080")) +
  geom_label(data = Snow.pumps, aes(x = x, y = y, label = label),
             size = 3, color = "blue", alpha = 0.8) +
  theme_void() +
  labs(
    title = "John Snow's Cholera Map",
    subtitle = "With contour lines showing densities of death",
    caption = "Data: HistData::Snow.deaths / Snow.pumps / Snow.streets"
  )
```

Adapted from `sandbox/Snow-density.R` (originally: John Russell,
[Challenge11.R](https://github.com/drjohnrussell/30DayChartChallenge/blob/main/2025/Challenge11.R)).

## Open invitations

Not attempted in this draft -- a literal "here's what's left" list:

- **`Minard`, `Guerry`, `Playfair1824`, `Pollen`, `OldMaps`** -- named and cited above, not worked
  through in depth in this first draft.
- **`Perozzo`'s red trace-line coloring** -- see "Still open" above.
- **`Perozzo` animated (`gganimate`)** -- held for a later draft; would add `gganimate` (+ a
  renderer) to `Suggests`.
- **Every other dormant `sandbox/` script** not chosen for the live case study above --
  `Galton-PearsonLee.R`, `Mcdonnell-density.R`, `Pyx-ggbump.R`, `Virginis-plot.R`,
  `Arbuthnot-PieGlyph.R`, `Cholera-plots.R`, `Nightingale-graph.R`, `Guerry-map.R`,
  `OldMaps-plot-map.R` -- each linked from its dataset's docs but never executed/shown; see the
  planning doc for what each would need beyond current `Suggests`.
- **`Ebbinghaus`** -- a not-yet-imported dataset (`issues/TASKS.md`), unrelated to any graphic
  challenge but still genuinely unfinished package work.
- **Rennie's fuller Snow map recreation** -- needs external data files not present in this repo;
  linked from `Snow.Rd` but can't be re-run here.
- **`Langren1644`** -- a *third* independent re-vision exists in [this
  blog post](https://friendly.github.io/blog/posts/2026-07-van-langren/)'s "Bonus: Reproducing
  Langren's 1644 Graph" section, alongside the package's own long-standing ggplot2 example and
  `sandbox/Langren-graph.R`. Worth a short paragraph on the design trade-offs between the three,
  not developed here yet.

## Going further

- Friendly, M. & Wainer, H. (2021). *A History of Data Visualization and Graphic Communication*.
  Harvard University Press.
- Companion site: <https://friendly.github.io/HistDataVis/> -- R code "reconstructions of
  historical graphs" tied to each book chapter.
- Gallery of Data Visualization (GDV), referenced in Friendly (2002).
- <https://www.datavis.ca/gallery/re-minard.php> -- the re-visions gallery this vignette's
  taxonomy is drawn from.

## References

*(Plain-text for now; convert to a `references.bib` + `[@key]` citations before this ships -- see
the note at the top of this file.)*

- Friendly, M. (2002). Visions and Re-Visions of Charles Joseph Minard. *Journal of Educational
  and Behavioral Statistics*, 27(1), 31-51.
- Friendly, M. (2007). A.-M. Guerry's Moral Statistics of France: Challenges for Multivariable
  Spatial Analysis. *Statistical Science*, 22, 368-399.
- Zabell, S. (1976). Arbuthnot, Heberden and the Bills of Mortality (Technical Report No. 40).
  Department of Statistics: The University of Chicago.
- Zabell, S., & Wainer, H. (2002). A Small Hurrah for the Black Death. *Chance*, 15(4), 58-60.
